\thetitle

### 2020-07-08 writing/rational.tex
revision of section 5

Based on this idea, a metric for migratory paths is a concise way to
evaluate the effectiveness of a
rational programmer. The components of each program create a \emph{configuration
space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
of variants of the program, dubbed \emph{configurations}, that differ in their
type specifications. In the setting of migratory typing,\footnote{The method can
be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each configuration differs from another in the subset
of typed components. Formally, a program $\system$ is a set of components
$\set{\component}$, and a configuration $\conf$ is a subset of
$\set{\component}$ . These configurations of $\system$ are ordered by the subset
relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
elements.  The bottom configuration of $\lattice{\system}$ is always $\emptyset$
and the top element is $\system$ itself. In between are all the mix-typed
configurations.

With these definitions, the actions of the rational programmer map to an
ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The starting
configuration $\conf_0$ of a trail---also referred to the \emph{root} (of the
trail)---is the configuration that the programmer attempts to debug. Any other
configuration $\conf_{i+1}$ in the trail differs from its predecessor
$\conf_{i}$ in that the untyped component that $\conf_i$ blames under the
various semantics, dubbed $\blame{\system, \conf_i}$, becomes typed in
$\conf_{i+1}$.\footnote{Blame can result either directly from a run time type
check or, in the absence of blame, from an exception that we track back to the
component that causes it with the help of the stack trace.}

@@

In natural Typed Racket, rational programmers rely on blame when they
attempt to debug a mix-typed programmer; they add type specifications to
the component blame points. In terms of configurations and trails, this
translates to extending a trail with a configuration that adds to the
last configuration of the trail so far the component the latter configuration
blames.  We call trails constructed in this manner \emph{natural blame
trails} and they capture the essence of the natural rational programmer:

\begin{quote}
\it 
A natural blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
  $\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

Natural rational programmers migrate faulty programs along natural blame
trails in order to locate bugs. Specifically, assuming that the types 
they add during migration are correct, when a buggy component migrates
from untyped to typed, the typechecker detects the impedance mismatch
between the broken code of the component and its types and signals a type
error. Hence, trails that end at an ill-typed configuration successfully 
pinpoint the location of bugs:
In formal terms:
\begin{quote}
\it 
A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful}  iff its last configuration $\conf_n$ does not type check.
  \end{quote}
  In juxtaposition:
\begin{quote}
\it 
A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{failing}  iff $\conf_n$ type checks and the trail cannot be extended
  further. 
 \end{quote}
\noindent 
That is failing natural blame trails are those that have not reached a configuration that reveals
the bug at compile time and their last configuration does not produce
blame. Thus the rational programmer receives no hint from the gradual type
system about how to further migrate the faulty program in search for the
bug.

@@

These definitions help a rational programmer gauge whether it pays off to heed blame
when debugging \system. 
However, they do not answer whether blame is a critical piece of 
the rational programmer's process.  For instance, 
typing a random component   could be equally useful as typing the blamed one. 
To account for this situation, we devise a new
mode of the natural rational programmer that 
 follows a random migration process:
\begin{quote}
\it 
A natural random trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
that is not in $\conf_i$.
\end{quote}
\indent
With blame and random natural trails, we can ask whether 
blame adds value over random migration: 
\begin{quote}
\it
  Blame adds value to natural for a program $\system$ over pure chance iff 
  no natural blame trails in  $\lattice{\system}$ are failing while 
 there exists at least one random natural trail that is
  failing.
\end{quote}

Even if the success of the natural rational programmer is not by pure
chance, it could well be because the natural semantics perform runtime type checks at
the right time. In other words, one rational programmer that 
heeds blame could be equally successful as one that only uses information from the exceptions of failed 
runtime type checks. Similar to before, in response we
define the exception mode of the rational programmer:
\begin{quote}
\it 
A natural exception trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
  where $\error{\system, \conf}$ denotes the component that triggers a
an exception of the underlying language.
\end{quote}
\noindent
We use the new mode to describe what it means for blame to add value over both exceptions and pure
chance:
\begin{quote}
\it
  Blame adds value to natural for a program $\system$ over pure chance and
  exceptions iff no natural blame trails in  $\lattice{\system}$ are failing while 
 there exist at least one random natural trail and one natural exception trail
 that are
  failing.
\end{quote}
\noindent
If blame does add value to natural for a program $\system$ 
over pure chance and exceptions then we can finally
deduce that, for debugging $\system$, blame is a valuable feature of the
natural semantics.

@@

 Since gradually typed languages with erasure semantics do not come with blame
assignment, a rational programmer can only hope that the underlying safety
checks and their exceptions are helpful.  Thus, adapting the corresponding
definitions from natural to erasure, we ask whether the effect of erasure
exception in debugging a program $\system$ goes beyond pure chance:

\begin{quote}
\it
  Exceptions adds value to erasure for a program $\system$ over pure chance
  iff no erasure exception trails in  $\lattice{\system}$ are failing while 
 there exists at least one random erasure trail that is failing.
\end{quote}
\noindent

@@

In contrast to erasure, transient does come with blame. As we discuss in
section~\ref{sec:landscape} though, distinctly from natural blame,
transient may blame more than one components. This peculiarity of 
transient raises the question of how the rational programmer should react
when the language produces a blame set. Our answer is that the rational
programmer has least two reasonable options.  The first option is the rational programmer types
all blamed components at once.  In the second option, the rational
programmer selects the component that is added to the blame set first and types
only that --- after all the types of this first component should be able
to  detect an impedance mismatch earlier in the evaluation of a program. 
As before, the two modes give rise to two different notions of trail:
\begin{quote}
\it 
A  transient all blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i = \mblame{\system, \conf_i}$
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics; 
\end{quote}
\noindent
and
\begin{quote}
\it 
A transient firstborn blame trail
is a sequence of configurations $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
  where $\first{\mblame{\system, \conf}}$ is the first component 
  the transient semantics add to the blame set for $\conf$.
\end{quote}
\noindent As for natural we use the transient all blame trails and 
transient firstborn  trails to define when  transient all blame and 
transient firstborn add value to transient over exceptions and pure
chance.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket and the 
mutants of section~\ref{sec:mutate}. In particular, our experiment examines 
four questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame is valuable for natural?

@@

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different trails/modes of the rational programmer. For example experimental
question $Q_1$ asks whether blame is valuable for natural and our experiment
uses the natural blame, exception and random trails to answer it.

@@

In detail, to answer $Q_1$ for a mutant, we first construct the mutant's configuration 
lattice. Then, we determine the configurations of the lattice that detect the fault in the
program. This can happen either because Typed Racket's type checker
outright rejects the configuration or because natural Typed Racket's runtime 
performs a type check that fails. In either
case,  each of these configurations is the root of a trail that we 
extend according the natural blame . 
If no configurations can be added to the trail\footnote{All trails
reach that point since the lattices are finite.}, we check if the
last configuration of the trail type checks or not. If it does the trail
is successful, otherwise it is failing. We repeat the same process for the
roots two more times; once following the natural exception rational programmer and 
once the natural random rational programmer. Finally, we use the
success/failure results for the trails to check if they meet the condition
for natural blame to add value to natural over exceptions and pure chance.
The process is analogues for the other questions and their corresponding
modes of the rational programmer.

@@

In addition to the value of blame, trails permit us to examine the effort a
rational programmer spends to locate a bug. As a metric for the effort, 
we use the number of components the rational programmer has to migrate 
from untyped to typed to find the bug. We can easily and uniformly calculate the value
of the metric for any successful trail in any mode of the rational
programmer (including transient blame all)  as the  size of the difference 
between the trail's root and its final configuration.

Programmer effort can help shed some light to the relative effectiveness of the
three gradual typing systems. For example, consider a configuration of the
lattice of one of our mutants. The configuration corresponds to a
 scenario that the rational programmer attempts to debug. If the
configuration results in blame both under, for instance, the natural and transient semantics,
then the natural blame rational programmer and the two transient blame
rational programmers (all and firstborn) can compete to see which one
reaches a configuration that doesn't type check with less effort. In
general, the effort of trails of different modes that have the same
root is a metric that allows us to compare the effectiveness of these
modes even if they correspond to different gradual typing systems.
Unfortunately, statistically speaking, such comparisons between two modes do not generalize 
beyond the successful trails that share their root in our experiments.
Still they allow us to quantify the difference between the modes of the rational
programmers and thus the different blame assignment strategies  for the specific
debugging scenarios we analyze in our experiment.

>>>>

Based on this idea, a metric for migratory paths is a concise way to
evaluate the effectiveness of a
rational programmer. The components of each program create a
\emph{debugging scenario
space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
of variants of the program, dubbed \emph{scenarios}, that differ in their
type specifications. In the setting of migratory typing,\footnote{The method can
be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each scenario differs from another in which
components are typed. Formally, a program $\system$ is a set of components
$\set{\component}$, and a scenario $\conf$ is a subset of
$\set{\component}$ . These scenarios of $\system$ are ordered by the subset
relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
elements.  The bottom scenario of $\lattice{\system}$ is always $\emptyset$
and the top one is $\system$ itself. In between are all the mix-typed
scenarios.

With these definitions, the actions of the rational programmer map to an
ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$. 
Intuitively, with each action the rational programmer  reduces the current
scenario to an ``easier'' one to debug. The easiest scenarios are those
that the type checker rejects outright. Assuming the types of the
program are correct, if the typechecker signals a type error, the error
pinpoints an impedance mismatch between the broken code of a component 
and its type. Hence the programmer knows which component to fix. 
The starting
scenario $\conf_0$ of a trail---also referred to as the \emph{root} (of the
trail)---is the scenario that the programmer attempts to debug initially. Any other
configuration $\conf_{i+1}$ in the trail differs from its predecessor
$\conf_{i}$ in that more 
components compared to  $\conf_{i}$ become typed. The rational programmer 
decides which are these components based on the feedback it receives from
the gradual typing system when running $\conf_{i}$.

\footnote{Blame can result either directly from a run time type
check or, in the absence of blame, from an exception that we track back to the
component that causes it with the help of the stack trace.}

@@

In natural Typed Racket, rational programmers rely on blame when they
attempt to debug a mix-typed programmer; they add type specifications to
the component blame points. In terms of scenarios and trails, this
translates to extending a trail with a new scenario that adds to the
latest scenario from the trail the component that latest scenario
blames.  We call trails constructed in this manner \emph{natural blame
trails} and they capture the essence of the natural rational programmer
:\footnote{We use trail and rational programmer interchangeably from now
on.}

\begin{quote}
\it 
A natural blame trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
  $\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

Natural rational programmers migrate faulty programs along natural blame
trails in order to locate bugs. As we mention above, assuming that the types 
they add during migration are correct, when the buggy component of a
program migrates from untyped to typed, the typechecker detects the issue
at compile time. Hence, trails that end at an ill-typed scenarios successfully 
pinpoint the location of bugs:
In formal terms:
\begin{quote}
\it 
A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful}  iff its last scenario $\conf_n$ does not type check.
  \end{quote}
  In juxtaposition:
\begin{quote}
\it 
A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{failing}  iff $\conf_n$ type checks and the trail cannot be extended
  further. 
 \end{quote}
\noindent 
That is failing natural blame trails are those that have not reached a scenario that reveals
the bug at compile time and their last scenario does not produce
blame. Thus the rational programmer receives no hints from the gradual type
system about how to further migrate the faulty program in search for the
bug.

@@

These definitions allow a rational programmer to gauge whether it pays off to heed blame
when debugging \system. 
However, they do not answer whether blame is a critical piece of 
the rational programmer's process.  For instance, 
typing a random component   could be equally useful as typing the blamed one. 
To account for this situation, we devise a new
mode of the natural rational programmer that 
 follows a random migration process:
\begin{quote}
\it 
A natural random trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
that is not in $\conf_i$.
\end{quote}
\indent
Similar to natural blame we determine how successful the random
migration is based on the success of natural random trails: 
\begin{quote}
\it
  Natural random is always useful for a program $\system$ iff 
  $\lattice{\system}$ contains no failing natural random trails.

   Natural random is sometimes useful for a program $\system$ iff 
  $\lattice{\system}$ contains successful natural random trails.

   Natural random is never useful for a program $\system$ iff 
  $\lattice{\system}$ contains no successful natural random trails.

Putting all these pieces together, we can ask whether 
blame adds value over random migration: 
\begin{quote}
\it
  Natural blame is more successful than natural random for $\system$ if 
  either (i) natural blame is always successful for $\system$ while natural random 
  is sometimes or never successful for $\system$, or (ii) 
  natural blame is sometimes successful for $\system$ while natural random 
  is never successful for $\system$.

  Natural blame adds value to natural for a program $\system$ over pure chance iff 
  natural blame is more successful than natural random.
\end{quote}

Even if the success of the natural rational programmer is not by pure
chance, it could well be because the natural semantics perform runtime type checks at
the right time. In other words, one rational programmer that 
heeds blame could be equally successful as one that only uses information from the exceptions of failed 
runtime type checks. Similar to before, in response we
define the exception mode of the rational programmer:
\begin{quote}
\it 
A natural exception trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
  where $\error{\system, \conf}$ denotes the component that triggers a
an exception of the underlying language.
\end{quote}
\noindent
We use the new mode to ask whether blame adds value over both exceptions and pure
chance in natural:
\begin{quote}
\it
  Natural blame adds value to natural for a program $\system$ over pure chance and
  natural exceptions iff natural blame is more successful than both natural
  random and natural exception.
\end{quote}
\noindent
If blame does add value to natural for a program $\system$ 
over pure chance and exceptions then we can finally
deduce that, for debugging $\system$, blame is a valuable feature of the
natural semantics.

@@

 Since gradually typed languages with erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, adapting the
 definitions related to exception and random trails from natural to
 erasure, we ask whether the effect of erasure exceptions in debugging a
 program $\system$ goes beyond that of pure chance:

\begin{quote}
\it
  Erasure exceptions add value to erasure for a program $\system$ over pure chance
  iff erasure exception is more successful than erasure random.
\end{quote}
\noindent

@@

In contrast to erasure, transient does come with blame. As we discuss in
section~\ref{sec:landscape} though, distinctly from natural blame,
transient may blame more than one components. This peculiarity of 
transient raises the question of how the rational programmer should react
when the language produces a blame set. Our answer is that the rational
programmer has at least two reasonable options.  The first option is the rational programmer types
all blamed components at once.  In the second option, the rational
programmer selects the component that is added to the blame set first and types
only that --- after all the types of this first component should be able
to  detect an impedance mismatch earlier in the evaluation of a program. 
As before, the two modes give rise to two different notions of trail:
\begin{quote}
\it 
A  transient all blame trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i = \mblame{\system, \conf_i}$
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics; 
\end{quote}
\noindent
and
\begin{quote}
\it 
A transient firstborn blame trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
  where $\first{\mblame{\system, \conf}}$ is the first component 
  the transient semantics add to the blame set for $\conf$.
\end{quote}
\noindent As for natural we use the transient all blame trail transient
firstborn, transient erasure and transient random trails to define when
transient all blame and transient firstborn add value to transient over
exceptions and pure chance: 
\begin{quote}
\it 
Transient all blame adds value to transient for a program $\system$ over
  pure chance and transient exceptions iff transient all blame 
  is more successful than both transient random and transient exception.

Transient firstborn blame adds value to transient for a program $\system$ over
  pure chance and transient exceptions iff transient all blame 
  is more successful than both transient random and transient exception.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In particular, our experimental
process collects data to answer four questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame valuable for natural?

@@

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example experimental
question $Q_1$ asks whether blame is valuable for natural and our experiment
uses the natural blame, exception and random trails to answer it.

@@

In detail, to answer $Q_1$ for a mutant, we have to 
calculate and log for each natural blame, natural exception and natural
random trail 
in the scenario lattice of the mutant whether it is successful or failing. 
Hence, we first construct the mutant's scenario
lattice and we determine the scenario of the lattice that can detect the fault in the
program. This can happen either because Typed Racket's type checker
outright rejects the scenario or because natural Typed Racket's runtime 
performs a type check that fails. In either
case,  each of these scenarios is the root of a trail that we attempt
extend according to the natural blame rational programmer. 
If no scenarios can be added to the trail\footnote{All trails
reach that point since the lattices are finite.}, we check if the
last scenario of the trail type checks or not. If it does this
natural blame trail
is successful, otherwise it is failing. We repeat the same process for the
roots two more times; once following the natural exception rational programmer and 
once the natural random rational programmer. Finally, based on the
collected data, we
calculate the success/failure results for the trails to determine whether
 natural blame is more successful than  natural exceptions and pure chance
 and we obtain the answer for $Q_1$.
The process is analogues for the other questions employing their corresponding
modes of the rational programmer.

@@

In addition to successful and failure information for the various trails,  
our experiment  records the number of components a rational programmer 
has to type along each successful trail ($\lvert \conf_n \setminus \conf_0
\rvert$). We use this number as the metric for the effort 
of the rational programmer to debug the root scenario of each successful trail.

Programmer effort can help shed some light to the relative effectiveness of the
three gradual typing systems. For example, consider a debugging scenario from  the
lattice of one of our mutants.  If the
scenario results in blame both under, for instance, the natural and transient semantics,
then the natural blame rational programmer and the two transient blame
rational programmers (all and firstborn) can compete to see which one
reaches an ill-typed scenarion with less effort. In
general, the effort of trails of different modes that have the same
root is a point of comparison for the effectiveness of these
modes even if the modes correspond to different gradual typing systems.
Unfortunately, statistically speaking, such comparisons between two modes do not generalize 
beyond the successful trails with common roots in our experiments.
Still they allow us to quantify the difference between the modes of the rational
programmers and thus the different blame assignment strategies  for the specific
debugging scenarios we analyze in our experiment.
